# Introduction

## Thanks 

Various people have helped in the process of putting together these materials (either knowingly, or unknowingly). Big thanks go to: 

- [Alexandra Chouldechova](https://www.andrew.cmu.edu/user/achoulde/)
- [Allison Horst](https://www.allisonhorst.com/)
- [Andrew Heiss](https://www.andrewheiss.com/)
- [Ben Baumer](https://www.smith.edu/academics/faculty/ben-baumer)
- [Benoit Monin](https://www.gsb.stanford.edu/faculty-research/faculty/benoit-monin)
- [Bodo Winter](https://bodowinter.com/)
- [David Lagnado](https://www.ucl.ac.uk/pals/research/experimental-psychology/person/david-lagnado/)
- [Ewart Thomas](https://profiles.stanford.edu/ewart-thomas)
- [Henrik Singmann](http://singmann.org/)
- [Julian Jara-Ettinger](https://psychology.yale.edu/people/julian-jara-ettinger)
- [Justin Gardner](https://profiles.stanford.edu/justin-gardner)
- [Kevin Smith](http://www.mit.edu/~k2smith/)
- [Lisa DeBruine](https://debruine.github.io/)
- [Maarten Speekenbrink](https://www.ucl.ac.uk/pals/research/experimental-psychology/person/maarten-speekenbrink/)
- [Matthew Kay](https://www.mjskay.com/)
- [Matthew Salganik](http://www.princeton.edu/~mjs3/)
- [Mika Braginsky](https://mikabr.io/) 
- [Mike Frank](https://web.stanford.edu/~mcfrank/) 
- [Mine Çetinkaya-Rundel](https://mine-cr.com/)
- [Nick C. Huntington-Klein](https://www.nickchk.com/)
- [Patrick Mair](https://psychology.fas.harvard.edu/people/patrick-mair)
- [Paul-Christian Bürkner](https://paul-buerkner.github.io/about/) 
- [Peter Cushner Mohanty](https://explorecourses.stanford.edu/instructor/pmohanty)
- [Richard McElreath](https://xcelab.net/rm/)
- [Russ Poldrack](https://profiles.stanford.edu/russell-poldrack) 
- [Stephen Dewitt](https://www.ucl.ac.uk/pals/research/experimental-psychology/person/stephen-dewitt/)
- [Solomon Kurz](https://solomonkurz.netlify.app/) 
- [Tom Hardwicke](https://tomhardwicke.netlify.app/)
- [Tristan Mahr](https://www.tjmahr.com/) 

Special thanks go to my teaching teams: 

- 2022: 
  - Ari Beller
  - Sarah Wu
  - Chengxu Zhuang 
- 2021: 
  - Andrew Nam 
  - Catherine Thomas 
  - Jon Walters
  - Dan Yamins
- 2020: 
  - Tyler Bonnen
  - Andrew Nam 
  - Jinxiao Zhang
- 2019:
  - Andrew Lampinen
  - Mona Rosenke 
  - Shao-Fang (Pam) Wang

## License and citation

This book is licensed under the [Creative Commons Zero v1.0 Universal license](https://github.com/psych252/psych252book/blob/master/LICENSE). If you find these materials helpful for your work, I'd appreciate you citing it: 

```
@book{gerstenberg2022methods,
  title = {Statistical methods for the behavioral and social sciences},
  author = {Tobias Gerstenberg},
  year = {2022},
  url = {https://psych252.github.io/psych252book/}
}
```

## List of R packages used in this book 

```{r, eval=FALSE, message=FALSE}
# RMarkdown 
library("knitr")      # markdown things 
library("kableExtra") # for nicely formatted tables

# Datasets 
library("gapminder")    # data available from Gapminder.org 
library("NHANES")       # data set 
library("titanic")      # titanic dataset

# Data manipulation
library("arrangements") # fast generators and iterators for permutations, combinations and partitions
library("magrittr")     # for wrangling
library("tidyverse")    # everything else

# Visualization
library("patchwork")    # making figure panels
library("ggpol")        # for making fancy boxplots
library("ggridges")     # for making joyplots 
library("gganimate")    # for making animations
library("GGally")       # for pairs plot
library("ggrepel")      # for labels in ggplots
library("corrr")        # for calculating correlations between many variables
library("corrplot")     # for plotting correlations
library("DiagrammeR")   # for drawing diagrams
library("ggeffects")    # for visualizing effects
library("bayesplot")    # for visualization of Bayesian model fits 

# Modeling 
library("afex")         # also for running ANOVAs
library("lme4")         # mixed effects models 
library("emmeans")      # comparing estimated marginal means 
library("broom.mixed")  # getting tidy mixed model summaries
library("janitor")      # cleaning variable names 
library("car")          # for running ANOVAs
library("rstanarm")     # for Bayesian models
library("greta")        # Bayesian models
library("tidybayes")    # tidying up results from Bayesian models
library("boot")         # bootstrapping
library("modelr")       # cross-validation and bootstrapping
library("mediation")    # for mediation and moderation analysis 
library("multilevel")   # Sobel test
library("extraDistr")   # additional probability distributions
library("effects")      # for showing effects in linear, generalized linear, and other models
library("brms")         # Bayesian regression
library("parameters")   # For extracting parameters 

# Misc 
library("tictoc")       # timing things
library("MASS")         # various useful functions (e.g. bootstrapped confidence intervals)
library("lsr")          # for computing effect size measures
library("extrafont")    # additional fonts
library("pwr")          # for power calculations
library("arrangements") # fast generators and iterators for permutations, combinations and partitions
library("stargazer")    # for regression tables
library("sjPlot")       # for regression tables
library("xtable")       # for tables
```

## Session info

```{r, echo=F}
sessionInfo()
```
