# Mixed effects model 

```{r echo=FALSE, message=FALSE}
library("knitr") #for markdown things 
library("broom") #for getting tidy model summaries
library("broom.mixed") #for getting tidy mixed model summaries
library("lme4") #for mixed effects models 
library("brms") #Bayesian regression
# library("patchwork") #for making figure panels
library("janitor") #for cleaning variable names
library("emmeans") #for comparing estimated marginal means 
library("pbkrtest") #model comparison for linear mixed effects models
library("patchwork") #figure panels
# library("boot") #for bootstrapping
# library("tictoc") #for timing things
# library("tidyboot") #for tidy bootstrapping (might not use this)
library("tidyverse") #for everything else
library("optimx") #for optimizing lmers
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

options(digits = 2) #only print two digits

theme_set(
  theme_classic()+
    theme(text = element_text(size = 20)) #text size
  )
```

## Todo 

- look into [ggeffects](https://strengejacke.github.io/ggeffects/) package 

## Resources 

### Online tutorials 
- [introduction to mixed models](https://ourcodingclub.github.io/2017/03/15/mixed-models.html)
- [tutorial on simulating a mixed model](https://aosmith.rbind.io/2018/04/23/simulate-simulate-part-2/)
- [how to build lmers](https://stats.stackexchange.com/questions/13166/rs-lmer-cheat-sheet)
- [lmer performance tips](https://cran.r-project.org/web/packages/lme4/vignettes/lmerperf.html)
- [lmer vignettes](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf)
- [data camp course](https://campus.datacamp.com/courses/hierarchical-and-mixed-effects-models/generalized-linear-mixed-effect-models?ex=9)

### Books 

- @gelman2006data


## Learning goals 

## Motiation for multi-level modeling 

- learning about treatment effects that vary: 
  - how does the effect of a treatment vary between groups within the population 

- using all the data to perform inferences for groups with small sample size: 
  - classical estimation is problematic for small sample sizes 
  - multilevel modeling allows estimate of group averages and group-level effects; a compromise between overly noisy within-group estimates and oversimplified regression that ignores group indicators 

- prediction: 
  - make predictions about individuals in new groups: group-effect is predicted from the group level model, observations predicted from the unit-level model

- analysis of structured data: 
  - some datasets have an inherent multilevel structure (e.g. students within schools, patients within hospitals, ...)
  - analysis should match the structure with which the data was generated as closely as possible 
  - multilevel models include indicators for clusters at all levels without falling prey to overfitting 

- more efficient inference for regression parameters: 
  - complete pooling --> differences between groups are ignored (one global regression)
    - suppresses variation 
  - no pooling --> data from different sources are analyzed separately (many separate individual regressions)
    - ignores information; yields very variable inferences 
  - partial pooling --> what comes out of multi-level model 
  
- including predictors at two different levels: 
  - classical regression cannot deal with nested predictors because of collinearity 
  
- getting the right standard error: accurately accounting for uncertainty in prediction and estimation 
  - account for correlation of the outcome between groups 
  
## Exploring the dependence assumption

```{r}
set.seed(0)
df.dependence = data_frame(
  participant = 1:20,
  condition1 = rnorm(20),
  condition2 = condition1 + rnorm(20, mean = 0.2, sd = 0.1)
) %>% 
  mutate(condition2shuffled = sample(condition2))

# df.dependence.shuffled = df.dependence %>% 
#   mutate(condition2 = sample(condition2)) #shuffle 

df.plot = df.dependence %>% 
  gather("condition", "value", -participant) %>% 
  mutate(condition = str_replace(condition, "condition", ""))

p1 = ggplot(data = df.plot %>% filter(condition != "2shuffled"), aes(x = condition, y = value))+
  geom_line(aes(group = participant), alpha = 0.3)+
  geom_point()+
  labs(title = "original",
       tag = "a)")
  

p2 = ggplot(data = df.plot %>% filter(condition != "2"), aes(x = condition, y = value))+
  geom_line(aes(group = participant), alpha = 0.3)+
  geom_point()+
  labs(title = "shuffled",
       tag = "b)")

p1 + p2 
  

```

```{r}
# separate the data sets 
df.original = df.dependence %>% 
  gather("condition", "value", -participant) %>% 
  mutate(condition = str_replace(condition, "condition", "")) %>% 
  filter(condition != "2shuffled")
  
df.shuffled = df.dependence %>% 
  gather("condition", "value", -participant) %>% 
  mutate(condition = str_replace(condition, "condition", "")) %>% 
  filter(condition != "2")
```

```{r}
# independent samples

# linear model 
lm(value ~ condition, data = df.original) %>% summary() 

# t.test
t.test(df.original$value[df.original$condition == "1"],
       df.original$value[df.original$condition == "2"],
       alternative = "two.sided",
       paired = F
       )
```

The mean difference between the conditions is very small 

```{r}
# dependent samples: original data set
fit.lmer1 = lmer(value ~ condition + (1 | participant), data = df.original)
fit.lmer2 = lmer(value ~ 1 + (1 | participant), data = df.original)

anova(fit.lmer1, fit.lmer2)

t.test(df.original$value[df.original$condition == "1"],
       df.original$value[df.original$condition == "2"],
       alternative = "two.sided",
       paired = T
       )
```


```{r}
# dependent samples: shuffled data set
fit.lmer1 = lmer(value ~ condition + (1 | participant), data = df.shuffled)
fit.lmer2 = lmer(value ~ 1 + (1 | participant), data = df.shuffled)

anova(fit.lmer1, fit.lmer2)

t.test(df.shuffled$value[df.shuffled$condition == "1"],
       df.shuffled$value[df.shuffled$condition == "2"],
       alternative = "two.sided",
       paired = T
       )
```



```{r}
# Bayesian test 

fit.brms1 = brm(value ~ condition + (1 | participant),
                data = df.original,
                file = "brms1")

# we use update() since we are fitting the same model to a new data set
fit.brms2 = update(fit.brms1, 
                   newdata = df.shuffled,
                   file = "brms2")

fit.brms1 %>% summary()

```

```{r}
df.plot = df.original %>% 
  bind_cols(fit.brms1 %>% 
              predict() %>% 
              as_data_frame() %>% 
              set_names(names(.) %>% tolower())) 

df.fit = fit.brms1 %>% 
  tidy()

ggplot(data = df.plot, aes(x = estimate, y = value, color = condition))+
  geom_abline(intercept = 0, slope = 1, linetype = 2, alpha = 0.2)+
  geom_linerange(aes(ymin = q2.5, ymax = q97.5), alpha = 0.3)+
  geom_point()

```



## Gelman book notes 

- key feature that distinguishes multilevel models from classical regression is in the explicit modeling of variation between groups

### Chapter 2: 

- probability distributions: 
  - distributions of data
  - distributions of parameter values 
  - distributions of error terms 
- 

## Tutorials

### Bodo Winter tutorial 2

- in mixed models, everything in the "fixed effects" part of the model works just like with linear models 
- the error terms ("random effects") are modeled more precisely --> random effects give structure to the error term 
- random effect for participant: 
  - different baseline value for each subject 
  - different items

- helpful visualizations: 
  - variability by subject
  - variability by item 

- in the past: 
  - averaging over subjects to get a by-item analysis 
  - averaging over items to get a by-subjects analysis
- mixed models take the full data into account (account for both item and subject variation)

```{r}
df.politeness = read.csv("data/politeness_data.csv")
  
df.politeness = df.politeness %>%
  mutate(gender = factor(gender, levels = c("F", "M"), labels = c("female", "male")),
         attitude = factor(attitude, levels = c("inf", "pol"), labels = c("informal", "polite"))) %>% 
  rename(pitch = frequency)

df.politeness %>% glimpse()
```

- Variation across subjects

```{r}
df.plot = df.politeness

ggplot(data = df.plot, aes(x = subject, y = pitch))+
  geom_point(alpha = 0.5, position = position_jitter(height = 0, width = 0.1))+
  stat_summary(fun.data = "mean_cl_boot", geom = "linerange", size = 1)+
  stat_summary(fun.y = "mean", geom = "point", shape = 21, fill = "red", color = "black", size = 3)+
  labs(y = "pitch (Hz)")
```

- Variation across scenarios

```{r}
df.plot = df.politeness %>% 
  mutate(scenario = as.factor(scenario))

ggplot(data = df.plot, aes(x = scenario, y = pitch))+
  geom_point(alpha = 0.5, position = position_jitter(height = 0, width = 0.1))+
  stat_summary(fun.data = "mean_cl_boot", geom = "linerange", size = 1)+
  stat_summary(fun.y = "mean", geom = "point", shape = 21, fill = "red", color = "black", size = 3)+
  labs(y = "pitch (Hz)")
```



```{r}
df.plot = df.politeness

ggplot(data = df.plot, aes(x = attitude, y = pitch))+
  geom_point(alpha = 0.5, position = position_jitter(height = 0, width = 0.1),
             aes(color = subject))+
  stat_summary(fun.data = "mean_cl_boot", geom = "linerange", size = 1)+
  stat_summary(fun.y = "mean", geom = "point", shape = 21, fill = "red", color = "black", size = 3)+
  facet_grid(~gender)+
  labs(y = "pitch (Hz)")
```

Simple linear model:

```{r}
fit.lm = lm(pitch ~ attitude, data = df.politeness)
fit.lm %>% summary()
```

Model interpretation 

Residuals: 

```{r}
df.plot = fit.lm %>% 
  augment() %>% 
  clean_names() %>% 
  left_join(df.politeness,
            by = c("pitch", "gender", "attitude"))
  

ggplot(data = df.plot, aes(x = fitted, y = resid, color = subject))+
  geom_point()+
  geom_smooth(method = "lm", aes(group = 1), show.legend = F)
```

Linear mixed effects model

```{r}
fit.lmer = lmer(pitch ~ attitude + (1 | subject) + (1 | scenario), data = df.politeness)
fit.lmer %>% summary()
```




```{r}
fit.lmer = lmer(pitch ~ attitude + gender + (1 | subject) + (1 | scenario), data = df.politeness)
fit.lmer %>% summary()
```

- variation for subject error term dropped (since much of the variation is explained by gender)

get p-values via the likelihood ratio test
- likelihood = probability of seeing the data given the model
- compare likelihood of two models: one with the term of interest, and one without
- for likelihood tests, set `REML=F` 

```{r}
fit.simple = lmer(pitch ~ gender + (1 | subject) + (1 | scenario), data = df.politeness, REML = F)
fit.complex = lmer(pitch ~ attitude + gender + (1 | subject) + (1 | scenario), data = df.politeness, REML = F)
fit.interaction = lmer(pitch ~ attitude * gender + (1 | subject) + (1 | scenario), data = df.politeness, REML = F)

anova(fit.simple, fit.complex)
anova(fit.complex, fit.interaction)
```

model parameters: 

```{r}
fit.complex %>% coef()
```

```{r}
df.plot = df.politeness

ggplot(data = df.plot, aes(x = attitude, y = pitch, group = subject, fill = subject))+
  geom_point(alpha = 0.5, position = position_jitterdodge(jitter.height = 0, dodge.width = 0.9, jitter.width = 0.1),
             aes(color = subject))+
  stat_summary(fun.y = "mean", geom = "line", position = position_dodge(width = 0.9), aes(color = subject))+
  stat_summary(fun.data = "mean_cl_boot", geom = "linerange", size = 1, position = position_dodge(width = 0.9), alpha = 0.5)+
  stat_summary(fun.y = "mean", geom = "point", shape = 21, color = "black", size = 3, position = position_dodge(width = 0.9))+
  facet_grid(~gender)+
  labs(y = "pitch (Hz)")
```

```{r}
# fit.test = lm(pitch ~ attitude * gender, data = df.politeness)
# fit.test = lm(pitch ~ gender, data = df.politeness)
# fit.test = lm(pitch ~ attitude, data = df.politeness)
# fit.test = lm(pitch ~ gender + attitude, data = df.politeness)
# fit.test = lmer(pitch ~ attitude + gender + (1 + attitude | subject), data = df.politeness)
# fit.test = lmer(pitch ~ attitude + (1 | subject), data = df.politeness)
# fit.test = lmer(pitch ~ attitude + gender + (1 | subject), data = df.politeness)
# fit.test = lmer(pitch ~ attitude + (1 + attitude | subject), data = df.politeness)
fit.test = lmer(pitch ~ attitude + gender + (1 + attitude | subject) + (1 + attitude | scenario), data = df.politeness)

fit.test2 = lmer(pitch ~ gender + (1 + attitude | subject) + (1 + attitude | scenario), data = df.politeness)

anova(fit.test, fit.test2)

fit.test %>% coef()
fit.test %>% fixef()

fit.test %>% summary()

df.plot = df.politeness %>% 
  left_join(fit.test %>% 
  augment() %>% 
  clean_names())

# df.plot %>% 
#   na.omit() %>% 
#   summarize(r = cor(pitch, fitted))

df.fit = fit.test %>% 
  tidy()

ggplot(data = df.plot, aes(x = attitude, y = pitch, group = subject, fill = subject))+
  geom_point(alpha = 0.5, position = position_jitterdodge(jitter.height = 0, dodge.width = 0.9, jitter.width = 0.1),aes(color = subject))+
  stat_summary(fun.y = "mean", geom = "line", position = position_dodge(width = 0.9), aes(color = subject))+
  stat_summary(fun.y = "mean", geom = "line", position = position_dodge(width = 0.9), aes(color = subject, y = fitted), linetype = 2, size = 1)+
  stat_summary(fun.data = "mean_cl_boot", geom = "linerange", size = 1, position = position_dodge(width = 0.9), alpha = 0.5)+
  stat_summary(fun.y = "mean", geom = "point", shape = 21, color = "black", size = 3, position = position_dodge(width = 0.9))+
  facet_grid(~gender)+
  # stat_summary(fun.y = "mean", geom = "line", aes(color = subject, y = fitted), linetype = 2, size = 1)+
  labs(y = "pitch (Hz)")
```

- [how many parameters does my lmer have?](https://stats.stackexchange.com/questions/180230/number-of-parameters-in-mixed-model)
  - `summary()` shows the fit parameters 

- which random effects to include? 
  - use a full model if it works (and makes sense)
  - random intercepts-only models have a relatively high Type I error rate 
  - keep it maximal [@barr2013random-e]



### Dragons tutorial

```{r}
load("data/lmm_tutorial/dragons.RData")
head(dragons)

dragons = dragons %>%
  mutate(bodyLength2 = scale(bodyLength)[,])

basic.lm <- lm(testScore ~ bodyLength2, data = dragons)
summary(basic.lm)

ggplot(dragons, aes(x = bodyLength, y = testScore)) +
  geom_point() +
  geom_smooth(method = "lm") 

#residual plot 

df.plot = basic.lm %>% 
  augment() %>% 
  clean_names()
  
ggplot(data = df.plot, aes(x = fitted, y = resid))+
  geom_hline(yintercept = 0)+
  geom_point()+
  geom_smooth()

# qq plot 

ggplot(data = df.plot, aes(sample = resid))+
  geom_abline(intercept = 0, slope = 1, linetype = 2)+
  geom_qq()+
  geom_qq_line()

# distributions for different ranges 

ggplot(data = dragons, aes(x = mountainRange, y = testScore))+
  geom_point(alpha = 0.2, position = position_jitter(width = 0.1, height = 0))+
  stat_summary(fun.y = "mean", geom = "point", color = "black")+
  stat_summary(fun.data = "mean_cl_boot", geom = "linerange")+
  theme(axis.text.x = element_text(size = 10))


ggplot(dragons, aes(x = bodyLength, y = testScore, colour = mountainRange))+
  geom_point(size = 2)+
  theme(legend.position = "bottom")

# multiple regression

mountain.lm <- lm(testScore ~ bodyLength2 + mountainRange, data = dragons)
summary(mountain.lm)

# linear mixed effects model 

mixed.lmer <- lmer(testScore ~ bodyLength2 + (1|mountainRange), data = dragons)
summary(mixed.lmer)

#residual plot 

df.plot = mixed.lmer %>% 
  augment() %>% 
  clean_names()
  
ggplot(data = df.plot, aes(x = fitted, y = resid))+
  geom_hline(yintercept = 0)+
  geom_point()+
  geom_smooth()

# qq plot 

ggplot(data = df.plot, aes(sample = resid))+
  geom_abline(intercept = 0, slope = 1, linetype = 2)+
  geom_qq()+
  geom_qq_line()

# sites are nested within mountain ranges 
dragons = dragons %>%
  mutate(sample = str_c(mountainRange, site, sep = "_"))

mixed.lmer2 = lmer(testScore ~ bodyLength2 + (1|mountainRange) + (1|sample), data = dragons)  # the syntax stays the same, but now the nesting is taken into account
summary(mixed.lmer2)

ggplot(dragons, aes(x = bodyLength, y = testScore, colour = site)) +
  facet_wrap(~mountainRange, nrow=3) +
  geom_point() +
  theme_classic() +
  geom_line(data = cbind(dragons, pred = predict(mixed.lmer2)), aes(y = pred)) +
  theme(legend.position = "none")

mixed.lmer2 %>%
  tidy() %>%
  dwplot(style = "distribution",
    show_intercept = T,
         vline = geom_vline(xintercept = 0, linetype = 2))

mixed.lmer2 %>%
  tidy() %>%
  clean_names() %>% 
  ggplot(aes(x = term, y = estimate))+
  geom_point()+
  geom_linerange(aes(ymin = estimate-std_error, ymax = estimate +std_error))+
  coord_flip()

```

- proportion of variance in random effect compared to total variance gives a sense for how much of the remaining variance (after taking into account the fixed effects) is explained by the random effect 

- crossed and random nested random effects
- not all mixed models are hierarchical (e.g. there can be crossed random factors that don't represent a hierarchy)
- _crossed_: 
  - possibility that the same subject is observed under multiple values of the random factor (e.g. the same dragon in different mountain ranges)
- _nested_: 
  - factor appears only within one particular level of another factor 


#### Linear regression on the aggregate

- doesn't take into account the uncertainty of individual aggregate data points (some means might be based on fewer observations, and so they shouldn't affect the overall as much) 

#### Separate individual regressions 

- many parameters 
- inflated Type I error (of falsely rejecting the null)
- 

#### Mixed effects models 

- use all the data we have 
- account for correlations within our data 
- estimate fewer parameters (hence more robust)
- avoide problems with multiple comparisons 
- confidence intervals on parameters 

```{r}
mixed.lmer2 = lmer(testScore ~ bodyLength2 + (1|mountainRange) + (1|sample), data = dragons)
mixed.lmer2 %>% confint(method = "boot",
                        nsim = 500,
                        boot.type = "perc")

# could add confidence intervals to table ...
tmp = mixed.lmer2 %>% 
  tidy()
```


```{r}
mixed.brm = brm(testScore ~ bodyLength2 + (1|mountainRange) + (1|sample), data = dragons)
mixed.brm %>% 
  tidy()

```

## Example

```{r}
# not sure whether the generating model is legit here ... 
set.seed(2)
df.data = data_frame(
  participant = rep(1:20,2),
  individual = rep(rnorm(20, mean = 0, sd = 10), 2), #explore for different values of SD 
  condition = rep(c(1, 2), each = 20),
  fixed.mean = ifelse(test = condition == 1, yes = 3, no = 10),
  x = runif(40, -10, 10),
  y = x + rnorm(40, mean = fixed.mean, sd = 1) + rnorm(40, mean = individual, sd = 1)
) %>% 
  mutate(condition = as.factor(condition))

fit.lm = lm(y ~ condition + x, data = df.data)
fit.lmer = lmer(y ~ x + condition + (1 | participant), data = df.data)
# fit.lmer = lmer(y ~ x + condition + (1 | participant), data = df.data)

#calculate confidence intervals 
fit.lm %>% confint()
fit.lmer %>% confint()

#summaries
fit.lmer %>% summary()
fit.lm %>% summary()

#fitted with lmer() 
p1 = ggplot(data = fit.lmer %>% augment, aes(x = .fitted, y = y))+
  geom_point(aes(color = as.factor(condition)))+
  geom_smooth(method = lm, se = F)+
  labs(title = "lmer()")+
  theme(legend.position = "none")

#fitted with lm()
p2 = ggplot(data = fit.lm %>% augment, aes(x = .fitted, y = y))+
  geom_point(aes(color = as.factor(condition)))+
  geom_smooth(method = lm, se = F)+
  labs(title = "lm()")+
  theme(legend.position = "bottom")

p1 + p2
```

```{r}
#main effect of condition
ggplot(data = df.data, aes(x = condition, y = y))+
  geom_point(aes(color = condition))+
  geom_line(aes(group = participant), alpha = 0.1)

ggplot(data = df.data, aes(x = x, y = y, color = condition))+
  geom_point()+
  geom_smooth(method = lm, se = F)+
  geom_line(aes(group = participant), alpha = 0.1)

# tmp = df.data %>% 
#   select(participant, condition, x, y) %>% 
#   spread(condition, y)
```


```{r}
#model comparison 
set.seed(2)
df.data = data_frame(
  participant = rep(1:20,2),
  individual = rep(rnorm(20, mean = 0, sd = 10),2), #explore for different values of SD 
  condition = rep(c(1, 2), each = 20),
  fixed.mean = ifelse(test = condition == 1, yes = 3, no = 10),
  x = runif(40, -10, 10),
  y = x + rnorm(40, mean = fixed.mean, sd = 1) + rnorm(40, mean = individual, sd = 1)
) %>% 
  mutate(condition = as.factor(condition))
  
# lmer 
fit.lmer1 = lmer(y ~ x + condition + (1 | participant), data = df.data)
fit.lmer2 = lmer(y ~ x + (1 | participant), data = df.data)
# fit.lmer2 = lmer(y ~ x * condition + (1 | participant), data = df.data)
anova(fit.lmer1, fit.lmer2)

fit.lmer1 = lmer(y ~ x + condition + (1 | participant), data = df.data)
fit.lmer2 = lmer(y ~ condition + (1 | participant), data = df.data)
anova(fit.lmer1, fit.lmer2)

# #lm
# fit.lm1 = lm(y ~ x + condition, data = df.data)
# fit.lm2 = lm(y ~ x , data = df.data)
# anova(fit.lm1, fit.lm2)

# alternative way using the car package 
# library(car)
# Anova(fit.lmer1)


```

## A mix of predictors ... 

- talk about standardizing predictors --> to interpret estimates 
  - differences expressed in standard deviations


## Comparison of two dependent samples 

```{r}
set.seed(0)
df.dependent.wide = data_frame(
  participant = 1:20,
  type = sample(c(-2, 2), size = 20, replace = T),
  cond_1 = rnorm(20, mean = 8, sd = 2) + rnorm(20, mean = type, sd = 2),
  cond_2 = rnorm(20, mean = 7, sd = 2) + rnorm(20, mean = type, sd = 2)
) 

# paired samples t-test 
t.test(df.dependent.wide$cond_1, df.dependent.wide$cond_2, paired = T, alternative = "greater")


# make a long data frame 
df.dependent.long = df.dependent.wide %>% 
  gather("condition", "rating", contains("cond"))

# linear mixed effects model 
tmp = lmer(rating ~ condition + (1 | participant), data = df.dependent.long) #different intercept for each participant

tmp %>% coef()
tmp2 = tmp %>% tidy

tmp2 = tmp %>% augment

df.plot = df.dependent.wide %>% 
  left_join(tmp %>% 
           augment() %>% 
           select(participant, coef = .fitted, .fixed),
           by = "participant") %>% 
  gather("condition", "rating", contains("cond"))

df.model = tmp %>% 
  augment

ggplot(data = df.plot, aes(x = condition, y = rating, color = participant))+
  geom_line(data = df.model, aes(y = .fitted, group = participant))+
  geom_point()+
  stat_summary(data = df.model, 
               fun.y = "mean",
               aes(x = condition, y = .fixed),
               geom = "point", 
               shape = 21,
               color = "black",
               fill = "red",
               size = 4)

```

## Fixed and random effects 

[difference between fixed and random effects](https://stats.stackexchange.com/questions/4700/what-is-the-difference-between-fixed-effect-random-effect-and-mixed-effect-mode)

- what are we interested in? 
- what do we want to control for / take into account? 

- fixed effects: variables that we expect will have an effect on the dv 
- random effects/factors: grouping factors that we try to control for 
  - data for random effect is just a sample of all the possibilities 
  - often: random effects should have more than five levels
- overall goal: have the model make better estimates 
- need to use random effects whenever we have repeated observations of the same object

## Model selection

- guided by theory
- not just randomly plugging in and dropping variables
- likelihood ratio tests using `anova()` (ok for large sample sizes)
 - only works for comparing more complex and simpler model 
 - only works for fixed effects
 - use maximum likelihood method when comparing models with different fixed effects (so set `REML = FALSE` when aiming to compare models via `anova()`)
 - use ML to compare models but still report parameter estimates from the best REML model (ML underestimates variance of the random effects)
- AIC for mixed effects models: `library("AICcmodavg")`

## Pooling 

[pooling visualizations](https://www.tjmahr.com/plotting-partial-pooling-in-mixed-effects-models/)

```{r}
df.sleep = sleepstudy %>% 
  as_data_frame() %>% 
  mutate(Subject = as.character(Subject))

# Add two fake participants
df.sleep <- bind_rows(
  df.sleep,
  data_frame(Reaction = c(286, 288), Days = 0:1, Subject = "374"),
  data_frame(Reaction = 245, Days = 0, Subject = "373")) %>% 
  clean_names()

df.sleep

```

Visualize the data 

```{r}
ggplot(df.sleep) + 
  aes(x = days, y = reaction) + 
  # stat_smooth(method = "lm", se = FALSE, fullrange = T) +
  stat_smooth(method = "lm", se = FALSE) +
  # Put the points on top of lines
  geom_point() +
  facet_wrap(~subject) +
  labs(x = "Days of sleep deprivation", 
       y = "Average reaction time (ms)") + 
  # We also need to help the x-axis, so it doesn't 
  # create gridlines/ticks on 2.5 days
  scale_x_continuous(breaks = 0:4 * 2)
```

- no pooling: separate regression line for each participant --> none of the information from different participants is shared/pooled 

```{r}
# df.no_pooling = lmList(reaction ~ days | subject, data = df.sleep) %>% 
#   coef() %>% 
#   # Subject IDs are stored as row-names. Make them an explicit column
#   rownames_to_column("subject") %>% 
#   rename(intercept = `(Intercept)`) %>% 
#   add_column(model = "No pooling") %>% 
#   # Remove the participant who only had one data-point
#   filter(subject != "373")

#alternative way 
df.no_pooling = df.sleep %>% 
  group_by(subject) %>%
  do(model = lm(reaction ~ days, data = .) %>% tidy()) %>%
  unnest() %>% 
  complete(subject, term) %>% 
  ungroup() %>% 
  select(subject:estimate) %>% 
  spread(term, estimate) %>% 
  rename(intercept = `(Intercept)`,
         days_slope = days) %>% 
  mutate(model = "no pooling")

#or 
# df.no_pooling3 = df.sleep %>% 
#   group_by(subject) %>% 
#   nest() %>%
#   mutate(model = map(data, ~lm(reaction ~ days, data = .) %>% tidy())) %>%
#   unnest(model) %>% 
#   complete(subject, term) %>% 
#   ungroup()

```

```{r}
fit.complete = lm(reaction ~ days, df.sleep)

df.complete_pooling <- data_frame(
  subject = unique(df.sleep$subject),
  intercept = coef(fit.complete)[1], 
  days_slope = coef(fit.complete)[2],
  model = "complete pooling")

```

Combine data sets and visualize 

```{r}
df.models = df.no_pooling %>% 
  bind_rows(df.complete_pooling) %>% 
  left_join(df.sleep, by = "subject")

p = ggplot(df.models) + 
  aes(x = days, y = reaction) + 
  # Set the color mapping in this layer so the points don't get a color
  geom_abline(aes(intercept = intercept, slope = days_slope, color = model),
              size = .75) + 
  geom_point() +
  facet_wrap(~subject) +
  labs(x = "Days of sleep deprivation", 
       y = "Average reaction time (ms)")+
  scale_x_continuous(breaks = 0:4 * 2) + 
  # Fix the color palette 
  scale_color_brewer(palette = "Dark2") + 
  theme(legend.position = "top")
p
```

 Partial pooling
 
```{r}
fit.partial = lmer(reaction ~ 1 + days + (1 + days | subject), data = df.sleep)
# arm::display(fit.partial)
summary(fit.partial)
```

Exctract parameters and plot results 

```{r}
df.partial = coef(fit.partial)[['subject']] %>% 
  as_tibble() %>% 
  rownames_to_column("subject") %>% 
  rename(intercept = `(Intercept)`, days_slope = days) %>% 
  add_column(model = "partial pooling")


df.models = bind_rows(df.complete_pooling, df.no_pooling, df.partial) %>% 
  left_join(df.sleep, by = "subject")


p %+% df.models
```

selection of participants 
 
```{r}
df.selection <- df.models %>% 
  filter(subject %in% c("335", "350", "373", "374"))

p %+% df.selection
```

- partial pooling model pulls extreme estimates towards an overall average 

visualization of shrinkage: 

```{r}
df.fixed_partial = data_frame(
  model = "partial pooling (average)",
  intercept = fixef(fit.partial)[1],
  days_slope = fixef(fit.partial)[2]
)

df.complete = df.complete_pooling %>% 
  distinct(model, intercept, days_slope)

df.averages = df.fixed_partial %>% 
  bind_rows(df.complete)

df.plot = df.no_pooling %>% 
  bind_rows(df.partial)

ggplot(df.plot) + 
  aes(x = intercept, y = days_slope, color = model) + 
  geom_point(size = 2) + 
  geom_point(data = df.averages, size = 5) + 
  # Draw an arrow connecting the observations between models
  geom_path(aes(group = subject, color = NULL), 
            arrow = arrow(length = unit(.02, "npc"))) + 
  # Use ggrepel to jitter the labels away from the points
  ggrepel::geom_text_repel(
    aes(label = subject, color = NULL), 
    data = df.no_pooling) + 
  theme(legend.position = "bottom") + 
  ggtitle("Pooling of regression parameters") + 
  xlab("Intercept estimate") + 
  ylab("Slope estimate") + 
  scale_color_brewer(palette = "Dark2") 

```

- shrinkage stronger for: 
  - more extreme values 
  - when data is limited 

- model avoids overfitting by taming extreme estimates 

Bayesian version 

```{r}
b <- stan_glmer(
  reaction ~ days + (days | subject),
  family = gaussian(),
  data = df.sleep,
  prior = normal(0, 2),
  prior_intercept = normal(0, 5),
  prior_covariance = decov(regularization = 2),
  prior_aux = cauchy(0, 1))
```


```{r}
df_posterior <- b %>% 
  as.data.frame() %>% 
  as_tibble()

ggplot(df_posterior) + 
  aes(x = `(Intercept)`, y = `days`) + 
  # Calculate the density
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  ggtitle("Where's the average intercept and slope?") + 
  xlab("Estimate for average intercept") + 
  ylab("Estimate for average slope") +
  # Use the same coordinate limits as last plot
  # coord_cartesian(
  #   xlim = range(df_pulled$Intercept), 
  #   ylim = range(df_pulled$Slope_Days),
  #   expand = TRUE) + 
  guides(fill = "none")
```


### Complete pooling 

### No pooling 

### Partial pooling 

## Simulating data 

- [crossvalidated post](https://stats.stackexchange.com/questions/187981/how-to-simulate-data-to-demonstrate-mixed-effects-with-r-lme4)
- [stackoverflow post](https://stackoverflow.com/questions/51937986/simulate-data-for-mixed-effects-model-with-predefined-parameter)

## Dice data 

```{r}
# read in data 

df.dice = read_csv("data/dice_data(main data).csv")
df.dice.additional = read_csv("data/dice_data(additional data).csv")

# combine and structure data 
df.dice = df.dice %>% 
  clean_names() %>% 
  select(participant:experiment) %>% 
  left_join(df.dice.additional %>% 
              clean_names() %>% 
              select(participant:experiment)) %>% 
  mutate(gender = factor(gender, levels = 1:2, labels = c("female", "male"))) %>% 
  group_by(experiment, participant) %>% 
  mutate(round = 1:n()) %>% 
  ungroup() %>% 
  mutate(phase = ifelse(round <= 20, 1, 2))

df.dice = df.dice %>% 
  gather("index", "value", die_p1:resp_p3) %>%
  separate(col = index, into = c("index", "player"), sep = "_") %>% 
  spread(index, value) %>% 
  arrange(participant, round) %>% 
  select(experiment, participant, round, phase, result, player, die, roll, payoff, resp) %>% 
  arrange(experiment, participant, round)

```

```{r}
# mixed effects model 

df.data = df.dice %>%
  filter(phase == 2) %>% 
  mutate(die = as.factor(die))

# positive results
# df.data = df.dice %>%
#   filter(result == 1, phase == 2) %>%
#   mutate(die = as.factor(die))

# negative results
# df.data = df.dice %>% 
#   filter(result == 0, phase == 2) %>% 
#   mutate(die = as.factor(die))

# fit2 = lmer(resp ~ 1 + die + roll + (1 | participant), data = df.data)
# fit1 = lmer(resp ~ 1 + die * roll + (1 | participant), data = df.data)
fit2 = lmer(formula = resp ~ 1 + (die + roll) * result + (1 + (die + roll) * result | participant),
            data = df.data,
            control = lmerControl(optimizer="Nelder_Mead"))

# fit2 = lmer(resp ~ 1 + die + roll + (1 + die | participant), data = df.data)
# fit3 = lmer(resp ~ 1 + die + roll + (1 + die | participant), data = df.data)
# fit4 = lmer(resp ~ 1 + die + roll + (1 + roll | participant), data = df.data)
# fit5 = lmer(resp ~ 1 + die * roll + (1 + die * roll | participant), data = df.data)

# fit %>% summary()
# fit1 %>% summary()
fit2 %>% summary()
# fit5 %>% summary()

tmp = coef(fit2)[["participant"]] %>% 
  as_data_frame() %>% 
  clean_names()

# anova(fit2, fit3)
# anova(fit5, fit2)

#overall plot 
df.fit.overall = fixef(fit2) %>% 
  as_data_frame() %>% 
  rownames_to_column() %>% 
  spread(rowname, value) %>% 
  clean_names()

ggplot(data = df.data, aes(x = roll, y = resp, group = die, color = die, fill = die))+
  stat_summary(fun.data = "mean_cl_boot", geom = "ribbon", alpha = 0.1, color = NA)+
  stat_summary(fun.y = "mean", geom = "line", size = 1)+
  scale_x_continuous(breaks = 1:6, labels = 1:6)+
  scale_y_continuous(breaks = 0:10, labels = 0:10)+
  coord_cartesian(ylim = c(0, 10))+
  # stat_summary(fun.y = "mean", geom = "line", linetype = 3, data = df.tmp, aes(y = fitted))+
  geom_abline(intercept = df.fit.overall$intercept, slope = df.fit.overall$roll, color = "red", linetype = 2)+
  geom_abline(intercept = df.fit.overall$intercept + df.fit.overall$die2, slope = df.fit.overall$roll, color = "green", linetype = 2)+
  geom_abline(intercept = df.fit.overall$intercept + df.fit.overall$die3, slope = df.fit.overall$roll, color = "blue", linetype = 2)
  
# individual plots
df.fit.participant = coef(fit2)[["participant"]] %>%
# df.fit.participant = coef(fit5)[["participant"]] %>% 
  clean_names() %>% 
  gather("index", "value") %>% 
  mutate(index = factor(index, levels = c("intercept", "die2", "die3", "roll")))

# visualize parameters 
ggplot(data = df.fit.participant, aes(x = value))+
  stat_density(geom = "line")+
  facet_wrap(~index, scales = "free")+
  theme(strip.background = element_blank())

df.tmp = fit2 %>%
# df.tmp = fit5 %>% 
  augment() %>% 
  clean_names()

# residuals
ggplot(df.tmp, aes(x = fitted, y = resid))+
  geom_point(alpha = 0.1)+
  geom_smooth(method = "loess")

ggplot(data = df.data, aes(x = roll, y = resp, group = die, color = die, fill = die))+
  # stat_summary(fun.data = "mean_cl_boot", geom = "ribbon", alpha = 0.1, color = NA)+
  stat_summary(fun.y = "mean", geom = "line")+
  geom_line(data = df.tmp, aes(y = fitted), linetype = 2)+
  facet_wrap(~ participant, ncol = 9)+
  theme(axis.text.y = element_blank(),
        strip.text = element_text(size = 6))

```

```{r}
# Bayesian version 

df.data.positive = df.dice %>% 
              filter(experiment == 2, phase == 2, result == 1) %>% 
              mutate(die = as.factor(die))

df.data.negative = df.dice %>% 
                            filter(experiment == 2, phase == 2, result == 0) %>% 
                            mutate(die = as.factor(die))

fit.brm.positive = brm(formula = resp ~ 1 + die + roll + (1 + die + roll | participant),
            data = df.data.positive,
            file = "fit.brm.positive")

fit.brm.negative = update(fit.brm.positive, 
                          newdata = df.data.negative,
                          file = "fit.brm.negative")


plot(marginal_effects(fit.brm.positive,
                      spaghetti = T, nsamples = 200),
     points = T,
     point_args = c(alpha = 1/2, size = 1))

fit.brm.positive %>% 
  tidy()

fit.brm.positive %>% 
  posterior_summary()

# positive 
df.parameters.positive = coef(fit.brm.positive)[["participant"]] %>% 
  as_data_frame() %>% 
  mutate(participant = 1:nrow(.)) %>% 
  select(-contains("error")) %>% 
  set_names(str_replace_all(names(.), "Q2.5", "low")) %>% 
  set_names(str_replace_all(names(.), "Q97.5", "high")) %>% 
  clean_names() %>%
  mutate(roll_index = ifelse(sign(low_roll) == sign(high_roll), 1, 0),
         die_index = ifelse(sign(low_die2) == sign(high_die2) | 
                             sign(low_die3) == sign(high_die3), 1, 0),
         both_index = roll_index * die_index,
         category = ifelse(roll_index, "roll", "neither"),
         category = ifelse(die_index, "die", category),
         category = ifelse(both_index, "both", category)
         ) %>% 
  select(-contains("index")) %>% 
  gather("index", "value", -c(participant, category)) %>% 
  separate(index, into = c("index", "parameter")) %>% 
  spread(index, value) %>% 
  mutate(parameter = factor(parameter, levels = c("intercept", "die2", "die3", "roll")))

# positive outcome 
df.model.positive = fit.brm.positive %>%
  augment() %>% 
  clean_names() %>% 
  bind_cols(df.data.positive) %>% 
  left_join(df.parameters.positive %>% 
              distinct(participant, category), by = "participant") %>% 
  mutate(participant = str_c(category, "_", participant))

df.plot.positive = df.data.positive %>% 
  left_join(df.parameters.positive %>% 
              distinct(participant, category), by = "participant") %>% 
  mutate(participant = str_c(category, "_", participant))

p.positive = ggplot(data = df.plot.positive, aes(x = roll, y = resp, group = die, color = die, fill = die))+
  stat_summary(fun.data = "mean_cl_boot", geom = "ribbon", alpha = 0.1, color = NA)+
  stat_summary(fun.y = "mean", geom = "line")+
  geom_line(data = df.model.positive, aes(y = fitted), linetype = 2)+
  facet_wrap(~ participant, ncol = 9)+
  theme(axis.text.y = element_blank(),
        strip.text = element_text(size = 6),
        strip.background = element_blank())

# negative 
df.parameters.negative = coef(fit.brm.negative)[["participant"]] %>% 
  as_data_frame() %>% 
  mutate(participant = 1:nrow(.)) %>% 
  select(-contains("error")) %>% 
  set_names(str_replace_all(names(.), "Q2.5", "low")) %>% 
  set_names(str_replace_all(names(.), "Q97.5", "high")) %>% 
  clean_names() %>%
  mutate(roll_index = ifelse(sign(low_roll) == sign(high_roll), 1, 0),
         die_index = ifelse(sign(low_die2) == sign(high_die2) | 
                             sign(low_die3) == sign(high_die3), 1, 0),
         both_index = roll_index * die_index,
         category = ifelse(roll_index, "roll", "neither"),
         category = ifelse(die_index, "die", category),
         category = ifelse(both_index, "both", category)
         ) %>% 
  select(-contains("index")) %>% 
  gather("index", "value", -c(participant, category)) %>% 
  separate(index, into = c("index", "parameter")) %>% 
  spread(index, value) %>% 
  mutate(parameter = factor(parameter, levels = c("intercept", "die2", "die3", "roll")))

# negative outcome 
df.model.negative = fit.brm.negative %>%
  augment() %>% 
  clean_names() %>% 
  bind_cols(df.data.negative) %>% 
  left_join(df.parameters.negative %>% 
              distinct(participant, category), by = "participant") %>% 
  mutate(participant = str_c(category, "_", participant))

df.plot.negative = df.data.negative %>% 
  left_join(df.parameters.negative %>% 
              distinct(participant, category), by = "participant") %>% 
  mutate(participant = str_c(category, "_", participant))

p.negative = ggplot(data = df.plot.negative, aes(x = roll, y = resp, group = die, color = die, fill = die))+
  stat_summary(fun.data = "mean_cl_boot", geom = "ribbon", alpha = 0.1, color = NA)+
  stat_summary(fun.y = "mean", geom = "line")+
  geom_line(data = df.model.negative, aes(y = fitted), linetype = 2)+
  facet_wrap(~ participant, ncol = 9)+
  theme(axis.text.y = element_blank(),
        strip.text = element_text(size = 6),
        strip.background = element_blank())

p.positive + p.negative + plot_layout(ncol = 1)

```




## Reporting results 

### Tables 

- stargazer package 
- stargazer cheatsheet 

### Plots 

- [dotwhisker plots](https://cran.r-project.org/web/packages/dotwhisker/vignettes/dotwhisker-vignette.html)

### Text 

from Bodo Winter tutorial

- likelihood ratio test: 

“... politeness affected pitch (χ2(1)=11.62, p=0.00065), lowering it by about 19.7 Hz ± 5.6 (standard errors) ...”

"We used R (R Core Team, 2012) and lme4 (Bates, Maechler & Bolker, 2012) to perform a linear mixed effects analysis of the relationship between pitch and politeness. As fixed effects, we entered politeness and gender (without interaction term) into the model. As random effects, we had intercepts for subjects and items, as well as by-subject and by-item random slopes for the effect of politeness. Visual inspection of residual plots did not reveal any obvious deviations from homoscedasticity or normality. P-values were obtained by likelihood ratio tests of the full model with the effect in question against the model without the effect in question."




```{r}
print(sessionInfo(), locale = FALSE)
```



