# Reporting statistics 

```{r, message=FALSE, include=FALSE}
library("knitr")      # for knitting RMarkdown 
library("kableExtra") # for making nice tables
library("janitor")    # for cleaning column names
library("broom")      # for tidying up model fits
library("lme4")       # mixed effects models 
library("brms")       # Bayesian regression
library("modelr")     # cross-validation and bootstrapping
library("tidybayes")  # tidying up results from Bayesian models
library("tidyverse")  # for wrangling, plotting, etc. 

opts_chunk$set(
  comment = "",
  results = "hold",
  fig.show = "hold"
)

theme_set(
  theme_classic() + #set the theme 
    theme(text = element_text(size = 20)) #set the default text size
)
```

In this chapter, I'll give a few examples for how to report statistical analysis. 

## General advice

Here is some general advice first: 

1. Make good figures! 
2. Use statistical models to answer concrete research questions.
3. Illustrate the uncertainty in your statistical inferences. 
4. Report effect sizes. 

### Make good figures!

Chapters \@ref(visualization-1) and \@ref(visualization-2) go into how to make figures and also talk a little bit about what makes for a good figure. Personally, I like it when the figures give me a good sense for the actual data. For example, for an experimental study, I would like to get a good sense for the responses that participants gave in the different experimental conditions. 

Sometimes, papers just report the results of statistical tests, or only visually display estimates of the parameters in the model. I'm not a fan of that since, as we've learned, the parameters of the model are only useful in so far the model captures the data-generating process reasonably well. 
### Use statistical models to answer concrete research questions.

Ideally, we formulate our research questions as statistical models upfront and pre-register our planned analyses (e.g. as an RMarkdown script with a complete analysis based on simulated data). We can then organize the results section by going through the sequence of research questions. Each statistical analysis then provides an answer to a specific research question. 

### Illustrate the uncertainty in your statistical inferences. 

For frequentist statistics, we can calculate confidence intervals (e.g. using bootstrapping) and we should provide these intervals together with the point estimates of the model's predictors. 

For Bayesian statistics, we can calculate credible intervals based on the posterior over the model parameters. 

Our figures should also indicate the uncertainty that we have in our statistical inferences (e.g. by adding confidence bands, or by showing some samples from the posterior). 

### Report effect sizes.

Rather than just saying whether the results of a statistical test was significant or not, you should, where possible, provide a measure of the effect size. Chapter \@ref(power-analysis) gives an overview of commonly used measures of effect size. 

<!-- ## Some concrete examples 

In this section, I'll give a few concrete examples for how to report the results of statistical tests. Each example tries to implement the general advice mentioned above. I will discuss frequentist and Bayesian statistics separately. 

### Frequentist statistics 

#### Simple regression 

```{r, message=FALSE}
df.credit = read_csv("data/credit.csv") %>% 
  rename(index = X1) %>% 
  clean_names()
```

__Research question__: Does a person's credit card balance increase with their level of income? 



```{r, fig.cap="Relationship between income level and credit card balance"}
ggplot(data = df.credit,
       mapping = aes(x = income,
                     y = balance)) + 
  geom_smooth(method = "lm",
              color = "black") + 
  geom_point(alpha = 0.2) +
  coord_cartesian(xlim = c(0, max(df.credit$income))) + 
  labs(x = "Income in $1K per year",
       y = "Credit card balance in $",
       title = "Relationship between income level and credit card balance.",
       subtitle = "The error band indicates a 95% confidence interval.")
```

### Bayesian statistics  -->




