--- 
title: "Psych 252: Statistical Methods for Behavioral and Social Sciences"
author: "Tobias Gerstenberg"
date: "`r Sys.Date()`"
book_filename: "psych252"
language:
  ui:
    chapter_name: "Chapter "
delete_merged_file: true
output_dir: "docs"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: psych252/psych252book
description: "Course notes for Psych 252."
---

# Preface {-}

Placeholder



<!--chapter:end:index.Rmd-->

# Introduction

## Thanks 

Various people have helped in the process of putting together these materials (either knowingly, or unknowingly). Big thanks go to: 

- Alexandra Chouldechova
- Ben Baumer
- Benoit Monin
- Datacamp
- David Lagnado
- Ewart Thomas
- Henrik Singmann
- Julian Jara-Ettinger
- Kevin Smith
- Maarten Speekenbrink
- Matthew Kay
- Matthew Salganik
- Mika Braginsky 
- Mike Frank 
- Mine Çetinkaya-Rundel
- Patrick Mair
- Peter Cushner Mohanty
- Richard McElreath
- Russ Poldrack 
- Stephen Dewitt
- Tom Hardwicke
- Tristan Mahr

Special thanks also to my teaching assistants: 

- Andrew Lampinen
- Mona Rosenke 
- Shao-Fang (Pam) Wang

## List of R packages used in this book 

```{r, eval=FALSE, message=FALSE}
# RMarkdown 
library("knitr")        # markdown things 
library("kableExtra")   # for nicely formatted tables

# Datasets 
library("gapminder")    # data available from Gapminder.org 
library("NHANES")       # data set 
library("titanic")      # titanic dataset

# Data manipulation
library("arrangements") # fast generators and iterators for permutations, combinations and partitions
library("magrittr")     # for wrangling
library("tidyverse")    # everything else

# Visualization
library("patchwork")    # making figure panels
library("cowplot")      # making figure panels
library("ggpol")        # for making fancy boxplots
library("ggridges")     # for making joyplots 
library("gganimate")    # for making animations
library("GGally")       # for pairs plot
library("ggrepel")      # for labels in ggplots
library("corrr")        # for calculating correlations between many variables
library("corrplot")     # for plotting correlations
library("DiagrammeR")   # for drawing diagrams

# Modeling 
library("afex")         # also for running ANOVAs
library("lme4")         # mixed effects models 
library("emmeans")      # comparing estimated marginal means 
library("broom")        # getting tidy model summaries
library("broom.mixed")  # getting tidy mixed model summaries
library("janitor")      # cleaning variable names 
library("car")          # for running ANOVAs
library("rstanarm")     # for Bayesian models
library("greta")        # Bayesian models
library("tidybayes")    # tidying up results from Bayesian models
library("boot")         # bootstrapping
library("modelr")       # cross-validation and bootstrapping
library("mediation")    # for mediation and moderation analysis 
library("multilevel")   # Sobel test
library("extraDistr")   # additional probability distributions
library("effects")      # for showing effects in linear, generalized linear, and other models
library("brms")         # Bayesian regression

# Misc 
library("tictoc")       # timing things
library("MASS")         # various useful functions (e.g. bootstrapped confidence intervals)
library("lsr")          # for computing effect size measures
library("extrafont")    # additional fonts
library("pwr")          # for power calculations
```


```{r}
library("knitr")      # for knitting RMarkdown 
library("kableExtra") # for making nice tables
library("janitor")    # for cleaning column names
library("greta")      # for writing Bayesian models
library("tidybayes")  # tidying up results from Bayesian models
library("cowplot")    # for making figure panels
library("ggrepel")    # for labels in ggplots 
library("gganimate")  # for animations
library("extraDistr") # additional probability distributions
library("tidyverse")  # for wrangling, plotting, etc. 

df.poker = read_csv("data/poker.csv") %>% 
  mutate(skill = factor(skill,
                        levels = 1:2,
                        labels = c("expert", "average")),
         skill = fct_relevel(skill, "average", "expert"),
         hand = factor(hand,
                       levels = 1:3,
                       labels = c("bad", "neutral", "good")),
         limit = factor(limit,
                        levels = 1:2,
                        labels = c("fixed", "none")),
         participant = 1:n()) %>% 
  select(participant, everything())


fit.brm1 = brm(formula = balance ~ 1 + hand,
               data = df.poker,
               file = "cache/brm1")

fit.brm1 %>% summary()

# generate predictive samples 
df.predictive_samples = fit.brm1 %>% 
  posterior_samples() %>% 
  clean_names() %>% 
  select(contains("b_"), sigma) %>% 
  sample_n(size = 20) %>% 
  mutate(sample = 1:n()) %>% 
  group_by(sample) %>% 
  nest() %>% 
  mutate(bad = map(data, ~ .$b_intercept + rnorm(100, sd = .$sigma)),
         neutral = map(data, ~ .$b_intercept + .$b_handneutral + rnorm(100, sd = .$sigma)),
         good = map(data, ~ .$b_intercept + .$b_handgood + rnorm(100, sd = .$sigma))) %>% 
  unnest(bad, neutral, good)

# plot the results as an animation
df.predictive_samples %>% 
  gather("hand", "balance", -sample) %>% 
  mutate(hand = factor(hand, levels = c("bad", "neutral", "good"))) %>% 
  ggplot(mapping = aes(x = hand,
                       y = balance,
                       fill = hand)) + 
  geom_point(alpha = 0.2,
             position = position_jitter(height = 0, width = 0.1)) + 
  stat_summary(fun.data = "mean_cl_boot",
               geom = "linerange",
               size = 1) + 
  stat_summary(fun.y = "mean",
               geom = "point",
               shape = 21,
               size = 4) +
  labs(y = "final balance (in Euros)") + 
  scale_fill_manual(values = c("red", "orange", "green")) +
  theme(legend.position = "none") + 
  transition_manual(sample)

```


<!--chapter:end:01-introduction.Rmd-->


# Visualization 1

Placeholder


## Learning objectives 
## Load packages
## Why visualize data?
### How _not_ to visualize data
### How to make it better
## Setting up RStudio
## Getting help
## Data visualization using `ggplot2`
### Setting up a plot
### Setting the default plot theme
### Scatter plot
#### Practice plot 1
### Line plot
### Adding error bars
#### Order matters
#### Practice plot 2
### Grouping data
#### Practice plot 3
### Making facets
#### Practice plot 4
### Global, local, and setting `aes()`
## Additional resources
### Cheatsheets
### Data camp courses
### Books and chapters
### Misc

<!--chapter:end:02-visualization1.Rmd-->


# Visualization 2

Placeholder


## Learning objectives 
## Install and load packages, load data, set theme
## Overview of different plot types for different things 
### Proportions 
#### Stacked bar charts 
#### Pie charts 
### Comparisons 
#### Boxplots
#### Violin plots
#### Joy plots
#### Practice plot
### Relationships 
#### Scatter plots
#### Raster plots 
### Temporal data 
## Customizing plots 
### Changing the order of things
### Dealing with legends 
### Choosing good colors
### Customizing themes 
## Saving plots 
## Creating figure panels 
## Peeking behind the scenes 
## Making animations 
## Shiny apps 
## Defining snippets 
## Additional resources 
### Cheatsheets 
### Data camp courses 
### Books and chapters
### Misc

<!--chapter:end:03-visualization2.Rmd-->


# Data wrangling 1 

Placeholder


## Learning objectives 
## Install packages 
## Load packages 
## Some R basics 
### Modes 
### Data types
#### Vectors 
#### Matrix 
#### Array 
#### Data frame 
#### Lists 
### Operators
### Control flow 
#### if-then {#if-else}
#### for loop
#### while loop 
### Functions 
#### Some often used functions 
### The pipe operator `%>%` 
#### Practice 1 
## Looking at data
### `head()`
### `glimpse()`
### `distinct()`
### `count()`
### `datatable()`
### Other tools for taking a quick look at data 
#### `vis_dat()`
#### `skim()`
#### `dfSummary()`
### A quick note on naming things 
## Wrangling data 
### filter() 
### rename() 
### select() 
### mutate() 
#### mutate_at()
#### mutate_all()
#### mutate_if()
### arrange() 
### Practice 2 
## Additional resources 
### Cheatsheets 
### Data camp courses
### Books and chapters

<!--chapter:end:04-data_wrangling1.Rmd-->


# Data wrangling 2 

Placeholder


## Learning objectives 
## Load packages 
## Wrangling data (continued)
### Summarizing data 
#### Practice 1 
### Reshaping data 
#### Practice 2 
### Joining multiple data frames 
#### Practice 3
### Dealing with missing data 
## Reading in data 
### csv
### RData 
## Saving data 
### csv 
### RData 
## Additional resources 
### Cheatsheets 
### Data camp courses 
### Books and chapters
### Tutorials 

<!--chapter:end:05-data_wrangling2.Rmd-->


# Probability and causality

Placeholder


## Load packages 
## Counting 
## Flipping a coin many times 
## Clue guide to probability 
### Conditional probability 
### Law of total probability
## Probability operations 
## Bayesian reasoning example
## Bayesian networks 
### Sprinkler example
## Additional resources 
### Cheatsheets 
### Books and chapters
### Misc 

<!--chapter:end:06-probability.Rmd-->


# Simulation 1 

Placeholder


## Load packages and set plotting theme  
## Working with distributions 
### Plotting distributions 
### Sampling from distributions 
### Cumulative probability distribution
### Inverse cumulative distribution 
### Computing probabilities 
#### Via probability distributions
#### Via sampling 
## Bayesian inference with the normal distribution
## Working with samples
### Understanding `density()`
### The `quantile()` function
## Comparing probability distributions 
## Additional resources 
### Cheatsheets 
### Datacamp 

<!--chapter:end:07-simulation1.Rmd-->


# Simulation 2 

Placeholder


## Load packages and set plotting theme  
## The central limit theorem 
### Population distribution 
### Distribution of a single sample 
### The sampling distribution
#### Bootstrapping a sampling distribution
### Exploring the CLT 
#### Distribution of height 
#### Testing the limits 
## Understanding p-values 
### Permutation test 
## Confidence intervals 
## Additional resources 
### Datacamp 

<!--chapter:end:08-simulation2.Rmd-->


# Modeling data

Placeholder


## Load packages and set plotting theme  
## Things that came up in class 
### Calculating RMSE using `magrittr` verbs
### Relationship between probability and likelihood 
## Modeling data 
### Simplicity vs. accuracy trade-off 
### Error definitions and best estimators
### Sampling distributions for median and mean 
## Hypothesis testing: "One-sample t-test" 
## Additional resources 
### Reading 
### Datacamp 

<!--chapter:end:09-modeling_data.Rmd-->


# Linear model 1

Placeholder


## Load packages and set plotting theme  
## Things that came up in class 
### Building a sampling distribution of PRE 
### Correlation 
#### Variance 
#### Covariance 
#### Spearman's rank order correlation
## Regression 
### Define and fit the models
### Calculate the sum of squared errors of each model
## Credit example
## Additional resources 
### Datacamp 
### Misc 

<!--chapter:end:10-linear_model1.Rmd-->


# Linear model 2

Placeholder


## Load packages and set plotting theme  
## Load data sets 
## Things that came up in class
## Multiple continuous variables 
### Explore correlations 
#### Visualize correlations
##### Correlations with the dependent variable
##### All pairwise correlations
### Multipe regression
#### Visualization
#### Fitting, hypothesis testing, evaluation
#### Visualizing the model fits 
#### Interpreting the model fits
#### Standardizing the predictors
## One categorical variable
### Visualization of the model predictions
### Dummy coding 
### Reporting the results
## One continuous and one categorical variable
### Visualization of the model predictions
## Interactions
### Visualization
### Hypothesis test 
## Additional resources 
### Datacamp 
### Misc 

<!--chapter:end:11-linear_model2.Rmd-->


# Linear model 3

Placeholder


## Load packages and set plotting theme  
## Load data sets 
### One-way ANOVA
#### Visualization 
#### Model fitting 
#### Hypothesis test 
#### Visualize the model's predictions 
#### Dummy coding
#### Follow up questions
## Two-way ANOVA 
### Visualization
### Model fitting
### Interpreting interactions 
## Additional resources 
### Datacamp 
### Misc 

<!--chapter:end:12-linear_model3.Rmd-->


# Linear model 4

Placeholder


## Load packages and set plotting theme  
## Load data sets
## Variance decomposition in one-way ANOVA
### Calculate sums of squares
### Visualize model predictions 
#### Total variance
#### Model variance
#### Residual variance 
## Two-way ANOVA (linear model)
### Calculate sums of squares
### Visualize model predictions
#### `Skill` factor
## ANOVA with unbalanced design
## Two-way ANOVA (with interaction)
## Planned contrasts 
### Hypothetical data 
### Visualization
### Constrasts 
### Post-hoc tests

<!--chapter:end:13-linear_model4.Rmd-->


# Power analysis

Placeholder


## Load packages and set plotting theme  
## Load data sets 
## Decision-making 
## Effect sizes
### eta-squared and partial eta-squared
### Cohen's d
## Determining sample size 
### `pwr` package
### Simulation
## Additional resources 

<!--chapter:end:14-power_analysis.Rmd-->


# Bootstrapping

Placeholder


## Load packages and set plotting theme  
## What's wrong with parametric tests?
### T-tests on non-normal distributions
### Non-IID noise and linear models
### Density estimate conceptual plot
## Bootstrap resampling
### Demo
## Applications
### Bootstrap confidence intervals
### Bootstrap (& permutation) hypothesis tests

<!--chapter:end:15-bootstrapping.Rmd-->


# Model comparison

Placeholder


## Load packages and set plotting theme  
## Determining sample size 
### `pwr` package
### `map()`
### via simulation
#### simple example 
#### more advanced example
## Model comparison 
### Fitting vs. predicting
### F-test 
### Cross-validation 
#### Leave-one-out cross-validation 
#### k-fold cross-validation 
#### Monte Carlo cross-validation 
### Bootstrap 
### AIC and BIC 
#### log() is your friend 
## Additional resources 
### Cheatsheet 
### Datacamp course
### Reading 
### Misc 

<!--chapter:end:16-model_comparison.Rmd-->


# Linear mixed effects models 1

Placeholder


## Load packages and set plotting theme  
## Things that came up in class 
### Comparing t-test with F-test in `lm()`
#### Global F-test 
#### Test for individual predictors
## Dependence 
## Additional resources 
### Readings 

<!--chapter:end:17-linear_mixed_effects_models1.Rmd-->


# Linear mixed effects models 2

Placeholder


## Load packages and set plotting theme  
## Things that came up in class 
### Difference between `replicate()` and `map()`
## Simulating a linear mixed effects model 
## The effect of outliers 
## Different slopes 
## Simpson's paradox 
## Additional resources 
### Readings 

<!--chapter:end:18-linear_mixed_effects_models2.Rmd-->


# Linear mixed effects models 3

Placeholder


## Load packages and set plotting theme  
## Load data set 
## Things that came up in class 
### One-tailed vs. two-tailed tests
#### t distribution
#### F distribution
### Mixtures of participants 
#### Ignoring mixture
#### Modeling mixture
#### Heterogeneity in variance
## Pooling and shrinkage 
### Complete pooling 
### No pooling 
### Partial pooling 
#### Random intercept and random slope
#### Only random intercepts 
#### Only random slopes 
### Compare results 
### Coefficients 
### Shrinkage 
## Bootstrapping 
### Linear model 
#### bootmer() function
## Getting p-values 
## Understanding the lmer() syntax 

<!--chapter:end:19-linear_mixed_effects_models3.Rmd-->


# Generalized linear model 

Placeholder


## Load packages and set plotting theme  
## Load data set 
## Logistic regression 
### Interpreting the parameters 
#### Binary predictor
#### Continuous predictor
#### Several predictors 
#### Using the "effects" package 
## Simulate a logistic regression
#### Calculate the model's likelihood 
## Testing hypotheses
## Logistic mixed effects model 
## Additional information 
### Datacamp 

<!--chapter:end:20-generalized_linear_model.Rmd-->


# Bayesian data analysis 1

Placeholder


## Load packages and set plotting theme  
## Things that came up 
### Bias in Cosyne 2019 conference admission? 

<!--chapter:end:21-bayesian_data_analysis1.Rmd-->


# Bayesian data analysis 2

Placeholder


## Load packages and set plotting theme  
## Doing Bayesian inference "by hand"
### Sequential updating based on the Beta distribution 
### Coin flip example 
### Bayesian inference by discretization
### Effect of the prior 
### Effect of the likelihood 
### Effect of the sample size  
## Distributions 
### Normal vs Student-t distribution
### Beta distributions
### Normal distributions 
### Distributions for non-negative parameters 
## Inference via sampling 
## Greta 
### Attitude data set 
### Frequentist analysis 
### Bayesian regression
#### Fit the model
#### Visualize the priors
#### Visualize the posteriors
#### Visualize model predictions 
#### Posterior predictive check 
#### Prior predictive check 

<!--chapter:end:22-bayesian_data_analysis2.Rmd-->


# Bayesian data analysis 3

Placeholder


## Load packages and set plotting theme  
## Load data set 
## Poker 
### Visualization
### Linear model 
### Bayesian model 
#### Visualize the posteriors 
#### Compute highest density intervals 
#### Posterior predictive check 
#### Test hypothesis
#### Bayes factor 
#### Full specification
##### Getting the priors
#### Inference diagnostics
##### When things go wrong 
## Dealing with heteroscedasticity 
## Ordinal regression 
## Additional resources 

<!--chapter:end:23-bayesian_data_analysis3.Rmd-->


# Mediation & Moderation

Placeholder


## Recommended reading 
## Load packages and set plotting theme  
## Mediation 
### Generate data 
### Method 1: Baron & Kenny’s (1986) indirect effect method
#### Total effect 
#### Path a 
#### Path b and c'
#### Interpretation
### Method 2: Sobel Test 
### Method 3: Bootstrapping
#### Interpretation 
### Method 4: Bayesian approach 
## Moderation 
### Generate data 
### Moderation analysis 
#### Visualize result 
## Additional resources 
### Books 
### Tutorials

<!--chapter:end:24-mediation_moderation.Rmd-->


# Cheatsheets

Placeholder


## Statistics 
## R 

<!--chapter:end:25-cheatsheets.Rmd-->

`r if (knitr::is_html_output()) '# References {-}'`

<!--chapter:end:26-references.Rmd-->

